smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world databases

### 2023-08-02 manuscript/manuscript.qmd
added objectives

In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms (see definitions box). These methods comprise standard procedures such as comparing distributions of baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed in both RWE and RCTs [@carroll2020; @wood2004; @harel2012].

@@

To overcome these shortcomings, we [@SentinelCI3] have recently developed and evaluated a principled approach combining a range of missing data diagnostics in an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized simulated missing data mechanisms well and provided helpful insights for the appropriate choice of analytic methods to handle the partially observed confounder data (e.g., missing data imputation).

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we here present and demonstrate the use of the `smdi` (structural missing data investigations) R package [@smdi]. To that end, we first provide an overview on the package's main functions including the underlying theoretical assumptions. We then demonstrate the practical application of the package by illustrating and end-to-end workflow with a hypothetical RWE study using a simulated oncology EHR dataset that comes bundled with the package.

@@

## Problem Formalization

As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data (RWD) are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., EHR). Hence, confounders and prognostic factors, which need to be balanced between treatment groups in RWE studies, are usually not available for all patients at all necessary time points [@toh2012a]. If the underlying mechanism for the missingness of such covariates is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to be missing, this can lead to bias in the resulting effect estimates for the studied treatments under common missing data approaches like complete case analysis or imputation [@groenwold2020]. Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for specific missing data approaches hold [@pedersen2017; @lee2013].

@@

...

@@

## Exemplary RWE Study and Data Generation

### Study Population

To showcase the practical implementation and suggested workflow when working with `smdi`, the package comes with a *de novo* simulated example EHR dataset that mimics a new-user cohort of `r formatC(nrow(smdi_data), format = "fg", big.mark = ",")` lung cancer patients initiating two comparable antineoplastic drug treatment regimens. The aim of the exemplary RWE study is to compare the comparative effectiveness of these two drug regimens in presence of confounding and three partially observed covariates, each following a different missingness mechanism. An overview of the dataset and baseline patient characteristics are provided in **`r table_cap("tbl2", display = "cite")`**.

### Simulation Design

The data generation process is described in detail in the Supplementary Methods. In brief, for this exemplary RWE study, we simulated a binary exposure (*Drug A versus Drug B*), a time-to-event outcome (*overall survival,* defined *as* as the time from treatment initiation until all-cause mortality or censoring) and 11 covariates, seven of which were true confounders (i.e., associated with both exposure and outcome), two were prognostic factors (i.e., associated only with the outcome but not with exposure) and two covariates which were neither associated with the exposure nor the outcome (**`r table_cap("tbl3", display = "cite")`**). We generated the survival times for the outcome assuming parametric event times following an exponential distribution and a truncated follow-up time of 5 years (Supplementary Table and Supplementary Figure ) [@simsurv]. The desired true treatment effect estimate was defined with a hazard ratio (HR) of 1.0. The strength of association between each covariate and the exposure and the outcome is displayed in Supplementary Table x and y, respectively.

### Missingness generation

To simulate missingness

>>>>

In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms (see definitions box). These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed [@carroll2020; @wood2004; @harel2012].

@@

To overcome these shortcomings, we [@SentinelCI3] have recently developed and evaluated a principled approach combining a range of missing data diagnostics in an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle the partially observed confounder data (e.g., missing data imputation).

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi]. The goal of `smdi` is to provide a toolbox of different diagnostics to scrutinize partially observed confounder data which had not been implemented elsewhere yet and complement other analytic R packages regularly used in RWE studies.

@@

@@

The smdi R package was written in R language (version 4.2.1)

@@
